---
title: "Making Graphs in R Using ggplot2: Part 1"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    df_print: default
runtime: shiny_prerendered
description: >
  Learn about the basic syntax of R.
---

## Overview

![](https://youtu.be/lbt4BH9Q82E)

In this session we will start looking at one of the most widely used, and almost universally useful, applications of the R programming languages - data visualisation. We will do this by learning specifically about the plotting library `ggplot2`. There are three parts to this workbook, as well as a series of exercises and questions. You can complete the workbooks at your own pace, across multiple sessions, although note that progress through each part of the workbook only saves if you are using the same device. However you can skip back and forth at any point within each part using the navigation menu.

### Installing the ggplot2 package
As mentioned in the video, this is not the default plotting library within R - but an additional package which needs to be installed seperately. We will learn more about installing packages a bit later in the course, but an important thing to remember is that when you are using RStudio or R, to make sure you have downloaded the package first,using `install.packages("ggplot2")`, and that you have loaded the package for the session, using `library(ggplot2)`. The package has already been pre-loaded into this workbook.


### Overview of ggplot2

`ggplot2` is also not the only additional plotting library that exists within R. There are many different ways of making plots using R, although these days `ggplot2` is the most commonly used. Particularly if you are looking at older resources online you may see plots being produced using the base R plotting libray, or sometimes in a different library called `lattice`. However, `ggplot2` is an obvious choice for new users of R to start with - since the code is designed to be consistent to make, theoretically, any sort of plot you could imagine, the default style is visually appealing but plots are also easily able to be customised in lots of ways. There is a trade off for this, since ggplot2 uses a lot of slightly strange terminology which may take a while to learn and get used to. 

'gg' stands for 'grammar of graphics', a term coined by the 1999 book by Leland Wilkinson. This is a system for defining graphs encouraging you to think about each of different components making up the graph, and providing specifications for those components. In this course we will focus on six of these components 'aesthetics','geometries','themes', 'facets', 'scales' and 'labels'. This specification makes you think clearly about exactly what you want to display, allows an extremely high degree of customisation both in 'graphical content' and in style, and makes it easy to provide visual consistency over lots of different graphs.

There are other components you can explore beyond this, but with these six components you will be able to make thousands of possible plots from your data. Both the ggplot2 syntax and the terminology can look confusing at first, but through practice once you understand the basic building blocks, you will be able to produce complex graphs very quickly and easily. 

As a reminder, in this workbook there are a mixture of code chunks that you can run without needing to modify any code, and code chunks which are prefaced with a **QUESTION**, where you will need to make changes to the code in the chunk. Solutions to these questions are embedded within the chunk.

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
tutorial_options(exercise.timelimit = 10)
Pulse<-readRDS("Pulse.RDS")

```


## 'Pulse' dataset 

The data we are using in this session is an extract of a survey conducted in the US in June 2018, as part of the Pulse of the Nation series of surveys. This survey was a representative sample of the American population. You can read more about the survey (and others from the same series) <a href="https://thepulseofthenation.com" target="_blank">here</a>. The data has been cleaned and edited slightly for use in this course.

The dataset contains an extract of 356 responses to 10 of the survey questions, and has been imported to R as a data frame called `Pulse`.

A summary of the columns in the dataset is below.

```{r, echo=FALSE,message=FALSE,warning=FALSE}


data.frame(Column=c("Gender","Age","AgeGrp","Race","Income","Education","PoliticalParty","PoliticalView","ApproveTrump","Attractiveness"),
           Description=c("What gender do you identify with?","What is your age?","Age range","What is your race?",
                         "About how much money do you make per year? (USD)","What is your highest level of education?",
                         "In politics today, do you consider yourself a Democrat, a Republican, or Independent?",
                         "Would you say you are liberal, conservative, or moderate?",
                         "Do you approve, disapprove, or neither approve nor disapprove of how Donald Trump is handling his job as president?",
                        "On a scale of 1-10, how physically attractive do you consider yourself?")) %>% kable()

```


Spend some time looking through the exploring the full dataset embedded below, to familiarise yourself with the columns and the type of data stored within each column. You may need to refer back to this data at times during this tutorial. Remember that R is case sensitive, so you will always have to refer to the variables in this dataset exactly as they are written in the data. There is a column in this data called "Gender" but there is no column in this data called "gender".

```{r,echo=FALSE}
DT::datatable(Pulse)
```

(You can use the arrow keys on your keyboard to scroll right in case the data table does not fit entirely on your screen)

## First Plot

To start off we are going to look at producing some plots with just one variable at a time. Let's start with a simple bar chart showing the frequency of different education levels within the dataset. The code has been written for you in the chunk below. Press "Run Code" in the chunk below to see the resulting plot.

```{r two_one,exercise=TRUE}
ggplot(data = Pulse,  aes(x = Education)) + 
  geom_bar()
```

Easy! Well, maybe not at first, since some of the commands we're using in that chunk look a bit strange. Let's start taking a look at each of the components within that code to understand what they all do.

## Data

Let's break down each of the different parts of this command.

With ggplot2, you always begin a graph with the function `ggplot()`. This function provides the information which will be passed to all subsequent functions making up the plot - about the data and the aesthetic mapping being used. 
And to make a plot you need some data, all ggplots must be made from an object of class data frame (or tibble) - we cannot plot anything which does not live inside of one of these data objects. 
So the first thing that we need to specify in the function is what data frame we are using. In this case the data frame we are using has been assigned the the name `Pulse`.

And we could just finish our code at this point and run this chunk of code, it is a valid and complete piece of code.

```{r data1,exercise=TRUE}
ggplot(data = Pulse)
```

But it is totally useless by itself! We have told R to make a plot from the `Pulse` data, but we have not told it what variables from the data we want to plot, or how we want to plot them. So we get a completely blank canvas returned to us.

This will lead us to specifying the two most important components of the ggplot - what variables we want to plot (the "aesthetic mapping") and how we want to plot them (the "geometries"). Every ggplot you will ever make will require you to specify as a minimum, the data, the aesthetics and the geometries. Everything else beyond that point is, to a certain degree, optional as there are default settings which are applied.


## Aesthetic Mapping

If we go back to the original code we saw the part of the code which said `aes(x = Education)`. 'aes' is the abbreviation for aesthetics. This is the word used to describe the different components of the plot onto which we can map different variables from our data.
When thinking of most straightforward graphs the obvious aesthetics are the x axis and y axis. But we will also learn more about other aesthetics later, as we can also use our data to define colours, size of points, shapes, labels and so on. There are some plot types, like bar charts, we only need to specify one aesthetic as a minimum. So we can map the column `Education` on to the x aesthetic of our graph in preparation for making a single bar chart.

Again, we can run just this line on its own:

```{r aes1,exercise=TRUE}
ggplot(data = Pulse, aes(x = Education))
```

You will see this is also a blank canvas, as before, except this time the x axis has been labeled and plotted based on the unique values of the column `Education` from our dataset `Pulse`. 

This whole line has now build up the page on which we will draw a plot. But we need to tell R exactly how it should draw a plot which we do by setting a geometry. 

## Geometries   

All geometry types are functions `geom_xxx` where the `xxx` is replaced by the shorthand for the plot type that you want to make. The geom function we want to use now for a bar chart is `geom_bar`

When we write the code for a ggplot, we think in terms of layering all of these different functions on top of each other. Each of the functions form a layer which are sequentially added to the plot.
The canvas we have just set is one layer; the bars will be another layer. The components are separated using a `+`, and it is usually conventional to start a new line for each new layer, as this makes your code easier to read.

```{r aes2,exercise=TRUE}
ggplot(data = Pulse,  aes(x = Education)) + 
  geom_bar()
```

Note that the `+` must be placed at the end of the line, not at the beginning of the new line. If you forget to put the plus at the end of the line, R will see a complete and valid piece of code and run this, and then ignore anything which comes afterwards (usually giving a warning or error while doing so).

The y axis for `geom_bar`, by default, becomes the frequency of each unique category within the variable mapped to our x aesthetic. So from the plot we can see that in our survey the most common education category was "college degree" with just over 100 respondents. 

**Question: Modify the code to make a bar chart of `Age` instead of `Education`**
```{r AgePlot,exercise=TRUE}
ggplot(data = Pulse,  aes(x = Education)) + 
  geom_bar()
```

```{r AgePlot-solution}
ggplot(data = Pulse,  aes(x = Age)) + 
  geom_bar()
```

## Plots with one continuous variable  

`Age`, unlike `Education`, is a continuous variable in our dataset. So a bar chart is probably not the best choice of plot when exploring our data. With this sort of data we might instead consider histograms, density plots or boxplots. These can be created easily by changing the geom from geom_bar to `geom_histogram` or `geom_density` or `geom_boxplot`, without really needing to modify our code in any other way.

```{r Agehist,exercise=TRUE}
ggplot(data = Pulse,  aes(x = Age)) + 
  geom_histogram()
```

It might be worth recapping the difference between bar charts and histograms if you are unsure of how these two types of plot vary, <a href="https://stattrek.com/statistics/charts/histogram.aspx" target="_blank">here</a>

`geom_density` also works in a similar way.
```{r Agedens,exercise=TRUE}
ggplot(data = Pulse,  aes(x = Age)) + 
  geom_density()
```
The density plot is a lot like a histogram, except it applies a smoother function so that you can try to understand the underlying distribution of the variable you are examining.

So let's try the same again with the boxplot:

```{r Agebox,exercise=TRUE,error=TRUE}
ggplot(data = Pulse,  aes(x = Age)) + 
  geom_boxplot()
```

It may be worth recapping the information presented within a boxplot, if it is not something you immediately remember,
<a href="https://stattrek.com/statistics/charts/boxplot.aspx" target="_blank">here</a>

Statisticians love boxplots. Although a single boxplot by itself is kind of boring - lots of boxplots side by side can be quite interesting and allow for comparisons. I wonder whether there is any relationship between political view and age?


With all of the plots we have made so far we have a choice - if we want the histogram, density plot or boxplot to be plotted horizontally we would specify just an x aesthetic. But with all of these geometries we could also have the same plot produced vertically by just using a y aesthetic.

```{r Agebox2,exercise=TRUE}
ggplot(data = Pulse,  aes(y = Age)) + 
  geom_boxplot()
```

In most cases it is personal preference as to whether you prefer to see the plots horizontally or vertically, and different people may have different preferences.

## Plots with more than variable

Using the same code we just used, we can add in a mapping of a categorical variable to our x aesthetic, to produce multiple boxplots side by side. With `geom_boxplot` our variable mapped to y must be a continuous numeric variable (like age or income) and our variable mapped to x must be a categorical (ordinal or nominal) variable (like political view or gender)

```{r Agebox3,exercise=TRUE}
ggplot(data = Pulse,  aes(y = Age,x=PoliticalView)) + 
  geom_boxplot()
```

*QUESTION: Try to interpret the results of this boxplot*

With `geom_boxplot` our variable mapped to y must be a continuous numeric variable (like age or income) and our variable mapped to `x` must be a categorical (ordinal or nominal) variable (like political view or gender). Placing a numeric variable, like `Attractiveness`, on the x axis for a boxplot, will not give an error, but the plot will just ignore that x variable and make a single box and a warning message will appear.

```{r Agebox4,exercise=TRUE}
ggplot(data = Pulse,  aes(y = Age,x=Attractiveness)) + 
  geom_boxplot()
```


**Question: Now modify the code below to produce some boxplots to investigate if there is a relationship between `Income` and `AgeGrp`. Make sure you assign the variables to the appropriate aesthetics**
```{r IncPol,exercise=TRUE}
ggplot(data = Pulse,  aes(y = Age,x=PoliticalView)) + 
  geom_boxplot()
```

```{r IncPol-solution}
ggplot(data = Pulse,  aes(y = Income,x=AgeGrp)) + 
  geom_boxplot()
```

### geom_scatter

We can now try and create a scatter plot, using geom_point(). However if we take the code from the first example and modify `geom_bar` to `geom_point` we find a different error.

```{r Agebox6,exercise=TRUE,error=TRUE}
ggplot(data = Pulse,  aes(x = Age)) + 
  geom_point()
```

This error is telling us that with have a "missing aesthetic: y". This is because we need to have both an x and a y variable to produce a scatter plot. Ideally for a scatter plot we need two numeric variables; so let's take a look at age on the x axis and attractiveness of respondent on the y axis.

```{r Agebox5,exercise=TRUE}
ggplot(data = Pulse,  aes(x = Age,y=Attractiveness)) + 
  geom_point()
```

Nothing immediately obvious is jumping out as a relationship between these variables!

As we have seen with `geom_point` and `geom_scatter`, different geometries will have different requirements for which 'aesthetics' it needs as a minimum requirement, and whether these variables should be numeric or categorical or something else. <a href="https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf" target="_blank">The ggplot2 cheat sheet</a> is a great resource to help work through this as is <a href="https://r-graphics.org/" target="_blank">the R graphics cookbook</a> 

Our plots are all a little basic so we should learn some more R so we can start to make them a bit more customised and impressive. In part two we will start to learn more about applying colours and scales to our plots.

## Appendix: Useful reference links  

R Graphics Cookbook: <a href="http://www.cookbook-r.com/Graphs/" target="_blank">http://www.cookbook-r.com/Graphs/</a> 

ggplot2 CheatSheet:<a href="https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf" target="_blank">https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf   </a>  

R Color Chart: <a href="http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf" target="_blank">http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf     </a>   

ggplot2 manuals (warning - technical and slightly pretentious. First line "ggplot2... embodies a deep philosophy of visualisation"): <a href="https://ggplot2.tidyverse.org/" target="_blank">https://ggplot2.tidyverse.org/     </a>     

Plotting with ggplot - Roger Peng videos:
<a href="https://www.youtube.com/watch?v=HeqHMM4ziXA" target="_blank">Part 1    </a>  
<a href="https://www.youtube.com/watch?v=n8kYa9vu1l8" target="_blank">Part 2 </a>  
